Understanding the Lomb-Scargle Periodogram

### 2017-01-04 PracticalLombScargle.tex
Add modifications section

A remarkable feature of Scargle's modified periodogram is that it is
{\it identical} to the result
obtained by fitting a simple sinusoid to the data at each frequency $f$, and
constructing a ``periodogram'' from the $\chi^2$ goodness-of-fit at each
frequency--an estimator which was considered in some depth by \citet{Lomb76}.
In this view, the $\tau$ shift defined in \eq{tau-def} serves to orthogonalize
the normal equations used in the least squares analysis.
In other words, the Lomb-Scargle periodogram essentially {\it assumes a
sinusoidal model} for the data, which is visualized in \fig{ls-model} for
the data we had seen in \fig{LINEAR-data}.

@@

Partly due to the deep connection between Fourier analysis and least-squares
analysis, the modified periodogram in \eq{lomb-scargle-periodogram}
has since become commonly referred to as the {\it Lomb-Scargle Periodogram}.
The Lomb-Scargle periodogram differs from the classical periodogram

@@

Suppose you have a set of $N$ observations over a time $T$ with an average
cadence of $\bar{\delta t} = N/T$.
From \eq{n-eval} we see that the number of frequencies evaluated is
$N_f \propto T f_{max} = N f_{max} / \bar{\delta t}$.
Holding constant the average survey cadence $\bar{\delta t}$ and individual
observation properties (which control $f_{max}$), we find that the number of
frequencies is simply proportional to the number of data points.
The computation of the Lomb-scargle periodogram in \eq{lomb-scargle-periodogram}
requires several sums over $N$ sinusoids at each of the
$N_f$ frequencies, and thus we see that the
na{\"i}ve scaling of the algorithm with the size of the dataset is
$\mathcal{O}(N^2)$, holding observing properties constant.
Due to the trigonometric functions involved, this turns out to be a rather
``expensive'' $\mathcal{O}(N^2)$, which makes the straightforward
implementation impractical for even modestly-sized datasets.

>>>>

A remarkable feature of Scargle's modified periodogram is that it is
{\it identical} to the result
obtained by fitting a model consisting of a simple sinusoid to the data at
each frequency $f$ and
constructing a ``periodogram'' from the $\chi^2$ goodness-of-fit at each
frequency--an estimator which was considered in some depth by \citet{Lomb76}.
In this view, the $\tau$ shift defined in \eq{tau-def} serves to orthogonalize
the normal equations used in the least squares analysis.

Partly due to the deep connection between Fourier analysis and least-squares
analysis, the modified periodogram in \eq{lomb-scargle-periodogram}
has since become commonly referred to as the {\it Lomb-Scargle Periodogram}.

\section{Least-Squares Extensions to Lomb-Scargle}
\sectlabel{lomb-scargle-extensions}

@@

The equivalence of the Fourier interpretation and least squares interpretation of the Lomb-Scargle periodogram allows for some interesting and useful extensions to the periodogram.
In the least squares interpretation of the periodogram, a sinusoidal model is proposed at each candidate frequency $f$:
\begin{equation}
  y_{model}(t;f) = A_f \sin(2 \pi f (t - \phi_f))
\end{equation}
where the amplitude $A_f$ and phase $\phi_f$ can vary as a function of frequency.
This is fit to the data in the standard least-squares sense, by constructing
the $\chi^2$ statistic at each frequency:
\begin{equation}
  \chi^2(f) \equiv \sum_n \big(y_n - y_{model}(t_n;f)\big)^2
  \eqlabel{chi2-simple}
\end{equation}
We can find the ``best'' model $\hat{y}(t;f)$ by minimizing $\chi^2(f)$ at
each freuqnecy with respect to $A_f$ and $\phi_f$;
we will denote this minimum value as $\hat{\chi}^2(f)$.
\citet{Scargle82} showed that with this setup, the Lomb-Scargle
periodogram from \eq{lomb-scargle-periodogram} can be equivalently written:
\begin{equation}
  P(f) = \frac{N}{2}\big[\hat{\chi}^2_0 - \hat{\chi}^2(f)\big]
  \eqlabel{lomb-scargle-chi2}
\end{equation}
where $\hat{\chi}^2_0$ is the reference model that assumes no variation in the data.
The key realization here is that the Lomb-Scargle periodogram essentially
{\it assumes a sinusoidal model} for the data; this is visualized in
\fig{ls-model} for the data we had seen in \fig{LINEAR-data}.
This immediately begs the question: can we compute a ``periodogram'' based
on more general forms of $y(t;f)$ to more effectively fit the data?

\subsection{Observational Noise}
\label{extensions-observational-noise}
Perhaps the most immediately obvious modification of the periodogram is to
change not the form of the model $y_{model}$, but the $\chi^2$ statistic itself
to account for observational errors in the data.
If there are Gaussian errors $\sigma_n$ on each observed $y_n$, we can re-write \eq{chi2-simple} in the following (standard) way:
\begin{equation}
  \chi^2(f) \equiv \sum_n \left(\frac{y_n - y_{model}(t_n;f)}{\sigma_n}\right)^2
  \eqlabel{chi2-with-errors}
\end{equation}
The periodogram constructed from this $\chi^2$ definition will more accurately
reflect the spectral power of noisy observations.
The effect of this modification on \eq{lomb-scargle-periodogram} is the addition
of a multiplicative weight $1/\sigma_n$ within each of the summations.
Early versions of this sort of modification appear in \citet{Gilliland87}
and \citet{Irwin89}; \citet{Zechmeister09} showed that such a modification
does not change the desirable statistical properties of the periodogram.

\subsection{Data Centering and the Floating Mean Periodogram}
\sectlabel{centering-data}

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{fig20_standard_vs_floatingmean}
  \caption{A comparison of the standard and floating mean periodograms for
    data with a frequency of 0.3 and a selection effect which removes
    faint observations.
    In this case the mean estimated from the observed data is not close to
    the true mean, which leads to the failure of the standard periodogram to
    recover the correct frequency. A floating mean model correctly recovers
    the true frequency of 0.3.
    \figlabel{standard-vs-floatingmean}}
\end{figure}

We have noted that the periodogram explored in \citep{Lomb76} and
\citep{Scargle82} is equivalent to computing the $\chi^2$ goodness-of-fit
of a sinusoid at each frequency.
Such a fit requires that the data be pre-centered; that is,
\begin{equation}
  \int_0^P g(t) dt = 0
\end{equation}fitting a sinusoid to the data at each
frrrequires the data to be pre-centered; i.e. the signal
must be modified so that

will only work correctly if the data is pre-centered so
that a sinusoid with no offset will fit the data.
In the past this was primarily accomplished by pre-centering the data using the
mean estimated from the data points themselves.
With full phase coverage of the observed signal, this approach is generally
robust, but due to selection effects and survey cadence, full coverage can
not always be guaranteed.
This can potentially lead to situations where the periodogram entirely misses
peaks of interest; an example of this is shown in \fig{standard-vs-floatingmean}.
This simulates noisy observations of a sinusoidal signal situation in which the faintest parts of the light curve fall below the detection threshold.
Using the standard periodogram, where the mean is estimated from the observations, leads to a periodogram that misses the true period of 0.3 days.
The floating-mean periodogram, on the other hand, fits the mean as part of the model and more accurately characterizes the periodic content of the partially-observed data.

\subsection{Fitting Higher-order Models}

\citet{Bretthorst88}

\todo{Use a multi-band periodogram on Kepler Data}

\todo{Show poor performance on eclipsing data}

@@

Suppose you have a set of $N$ observations over a time-span $T$ for an average
cadence of $\overline{\delta t} = N/T$.
From \eq{n-eval} we see that the number of frequencies evaluated is
$N_f \propto T f_{max} = N f_{max} / \overline{\delta t}$.
Holding constant the average survey cadence $\overline{\delta t}$ and individual
observation properties (which control $f_{max}$), we find that the number of
frequencies is directly proportional to the number of data points.
The computation of the Lomb-scargle periodogram in \eq{lomb-scargle-periodogram}
requires sums over $N$ sinusoids at each of the
$N_f$ frequencies, and thus we see that the
na{\"i}ve scaling of the algorithm with the size of the dataset is
$\mathcal{O}(N^2)$, when survey properties are held constant.
Due to the trigonometric functions involved, this turns out to be a rather
``expensive'' $\mathcal{O}(N^2)$, which makes the straightforward
implementation impractical for even modestly-sized datasets.
